Meta-Analysis of the 'Ironic' Effects of Intergroup Contact

### 2021-05-14 manuscript/manuscript.Rmd
submit

---
title: "Meta-Analysis of the 'Ironic' Effects of Intergroup Contact"
header: "'IRONIC' EFFECTS META-ANALYSIS"
keywords: intergroup contact, meta-analysis, perceived discrimination, collective action, policy support, social change, social injustice 
bibliography: references.bib
csl: apa.csl
abstract: 
  Critics have argued that intergroup contact, psychology’s most-researched paradigm for reducing prejudice, has the ‘ironic’ effect of stifling support for social change in disadvantaged groups. We conducted a preregistered meta-analytic test of this effect across `r with(dplyr::distinct(readRDS("../data/dl.rds"), sample, id, n), numform::f_comma(dplyr::n_distinct(id)))` studies with `r with(dplyr::distinct(readRDS("../data/dl.rds"), sample, id, n), numform::f_comma(length(sample)))` samples of `r with(dplyr::distinct(readRDS("../data/dl.rds"), sample, id, n), numform::f_comma(sum(n)))` disadvantaged-group members. As predicted, intergroup contact was, on average, associated with less perceived injustice (`r with(readRDS("../results/results_preregistered_analyses.rds"), paste0("$r = ", numform::f_num(median(r_mean[x_var == "ic" & y_var == "pi"]), digits = 2), "$"))`), collective action (`r with(readRDS("../results/results_preregistered_analyses.rds"), paste0("$r = ", numform::f_num(median(r_mean[x_var == "ic" & y_var == "ca"]), digits = 2), "$"))`), and support for reparative policies (`r with(readRDS("../results/results_preregistered_analyses.rds"), paste0("$r = ", numform::f_num(median(r_mean[x_var == "ic" & y_var == "ps"]), digits = 2), "$"))`) in disadvantaged groups. However, these associations were small, variable, and consistent with alterative explanations. Moderator analyses found that these associations were strongest in studies on short-term migration and post-colonial relations, that measured intergroup contact directly, and that involved adults. After controlling for the positive association of negative contact with support for social change, positive contact was not (negatively) associated with any outcome in studies that measured both predictors. We close by highlighting open questions about the relation between intergroup contact and social change.
statement: 
  As societies are becoming more diverse, there are more opportunities for members of different groups to come in contact with each other. Psychologists disagree about what this means for overcoming persistent inequality between advantaged and disadvantaged groups. Many argue that intergroup contact reduces advantaged-group members' prejudice and, thereby, prevents interpersonal discrimination. Reducing prejudice, however, is rarely enough to overcome institutional and structural discrimination. Instead, social change often results from disadvantaged-group members mobilizing against social injustice (for example, in the 2020 Black Lives Matter protests). Critics argue that positive intergroup contact could create a false sense of equality and, thereby, stifle disadvantaged-group members' opposition to injustice. We present the first systematic review and meta-analysis of the growing number of studies examining how intergroup contact relates to support for social change in disadvantaged groups. Our findings help us understand whether, when, and how intergroup contact might hinder, rather than help, social change.
output:
  pdf_document:
    latex_engine: pdflatex
    keep_tex: yes
    template: apa-template.tex
  word_document:
    reference_docx: word-template.docx
---

@@

It is ironic that psychology’s most-researched paradigm for reducing prejudice—intergroup contact—should hinder, rather than help, social change. Hundreds of studies have confirmed that intergroup contact is associated with less negative feelings toward outgroup members [for a meta-analysis, see @pettigrew_meta-analytic_2006]. Most of these studies have focused on advantaged-group members' contact experiences with disadvantaged-group members [@tropp_relationships_2005], guided by the idea that reducing advantaged-group members' prejudice will prevent interpersonal discrimination. Reducing prejudice, however, is rarely enough to overcome institutional and structural discrimination. Instead, as the 2020 Black Lives Matter protests have shown, social change often requires political mobilization by the disadvantaged. Critics have argued that, by fostering harmony and reducing conflict, intergroup contact might distract disadvantaged-group members from the injustice they face and undermine their motivation for social change [@dixon_beyond_2005; @reicher_rethinking_2007; @brown_strategic_2003]. By stifling disadvantaged-group members' opposition to injustice, intergroup contact might thus have the 'ironic' [@saguy_irony_2009], 'sedative' [@cakal_investigation_2011], or 'paradoxical' [@dixon_paradox_2010] effect of hindering social change.

Supporting this argument, initial studies provided evidence that, among disadvantaged-group members, intergroup contact reduces perceptions of injustice [@saguy_irony_2009], discourages collective action [@cakal_investigation_2011], and diminishes support for reparative policies [@dixon_intergroup_2007]. Others, however, have pointed to conflicting evidence [@poore_contact_2002] and argued that intergroup contact might instead render discrimination more salient and, thereby, increase disadvantaged-group members' support for social change [@pettigrew_recent_2011]. A decade after the initial studies, many more studies have examined the relationships between intergroup contact and support for social change in disadvantaged groups. We present a systematic review and meta-analysis of this growing literature. Specifically, we evaluate the evidence for three hypotheses about how intergroup contact might reduce support for social change.

Perceived injustice—that is, perceiving unjust group-based deprivation or discrimination against the disadvantaged ingroup—is a prerequisite to instigating social change [@van_zomeren_toward_2008]. Researchers have argued that positive contact with advantaged-group members could contradict perceptions of personal discrimination, undermine negative characterizations of the advantaged outgroup, and, thereby, make group discrimination appear less plausible [@dixon_paradox_2010]. Positive contact could further reduce perceived injustice by emphasizing commonalities over differences [@saguy_irony_2009] and by motivating disadvantaged-group members to adopt system-justifying ideologies [@sengupta_perpetuating_2013]. To the extent that it reduces perceived injustice, intergroup contact hinders social change.

Collective action—that is, any action directed at improving the conditions of the disadvantaged ingroup [@wright_responding_1990]—is crucial to achieving social change because advantaged-group members rarely surrender their advantages without sustained pressure from disadvantaged-group members [@blumer_race_1958; for a review, see @dixon_beyond_2012]. Researchers have argued that intergroup contact could discourage disadvantaged-group members from engaging in collective action by diminishing their perceived discrimination [@tropp_crossethnic_2012], by blurring boundaries between the disadvantaged ingroup and the advantaged outgroup [@saguy_irony_2009], by reducing negative attitudes toward the advantaged outgroup [@wright_struggle_2009], and by quelling anger about inequality [@tausch_how_2015] or toward the advantaged outgroup [@hayward_how_2018]. To the extent that it discourages collective action, intergroup contact hinders social change.

Reparative policies (e.g., affirmative action) can be effective means to reduce social inequality. Disadvantaged-group members, who stand to benefit from their implementation, are important advocates for these policies. Researchers have argued that intergroup contact could diminish disadvantaged-group members’ support for redistributive policies by creating sympathy for advantaged-group members [@dixon_intergroup_2007], by fostering positive outgroup attitudes and reducing attention to inequality [@saguy_irony_2009], and by increasing their endorsement of system-justifying ideologies [@sengupta_perpetuating_2013]. To the extent that it diminishes support for redistributive policies, intergroup contact hinders social change.

We conducted a preregistered meta-analysis of the evidence for the hypothesized negative effects of intergroup contact on perceived injustice, collective action, and policy support in disadvantaged groups. Our first objective was to systematically review the available evidence and to evaluate its strengths and limitations. Our second objective was to synthesize the available evidence to establish the direction and magnitude of the relationships between intergroup contact and the three outcomes. Doing so allowed us, first, to determine whether the available evidence supports a negative or positive [see @pettigrew_recent_2011] relationship between contact and support for social change and, second, to estimate the strength of this relationship. Our third objective was to explore the variability of this relationship across study settings, designs, and other potential moderators. For example, we examined whether any effects were confined to Western, Educated, Industrialized, Rich, Democratic (WEIRD) settings, the cultural context of most psychological research [@henrich_weirdest_2010]. Our fourth objective was to assess how robust the available evidence was to publication bias.

In addition, we collated the evidence for two alternative accounts of this relationship. First, we considered studies that measured both ingroup and outgroup contact as predictors of the three outcomes. @sengupta_ingroup_2015 argued that contact with other members of the disadvantaged ingroup would increase political mobilization in disadvantaged groups. As outgroup contact places limits on how much ingroup contact disadvantaged-group members can have [@pfister_contact_2020], the mobilizing effect of ingroup contact is an alternative explanation for negative associations between outgroup contact and support for social change. Second, we considered studies that measured both negative and positive (outgroup) contact as predictors of the three outcomes. Researchers [@hayward_how_2018; @reimer_intergroup_2017] found negative contact with advantaged-group members to be associated with greater perceived discrimination and stronger collective action intentions. @reimer_intergroup_2017 argued that, to the extent disadvantaged-group members who have more positive contact also tend to have less negative contact, the mobilizing effect of negative contact is an alternative explanation for negative associations between outgroup contact and support for social change.

@@

We included studies if they measured or manipulated the quantity of, quality of, and opportunity for contact with members of outgroups that are relatively advantaged compared to the participants’ ingroup. We considered opportunity for contact, as it is a potential precursor to and proxy for face-to-face contact, but not imagined or indirect contact. Similarly, we included studies that measured intergroup contact indirectly by, for example, asking what proportion of someone’s friends were *not* from the participants’ ingroup.

@@

***Perceived Injustice.*** We included studies that measured perceptions that one is discriminated against because of one’s group membership, that one’s group faces discrimination in society, that one’s group is relatively deprived compared to other groups, or that the deprivation and/or discrimination faced by one’s group is unjust and illegitimate. We also considered expectations of fair treatment as the reverse of perceived injustice. We included studies that measured personal and/or group discrimination or a mixture of both.

@@

Bayesian inference involves choosing a likelihood function and prior distributions. A likelihood function links the observed data to one or more model parameters and states how likely the observed data are given different values of said model parameters. Prior distributions state how plausible different values of said model parameters are before considering the observed data. Bayesian inference applies Bayes’ theorem to update prior distributions in light of the observed data to produce posterior distributions. Other than *p*-values and confidence intervals, the resulting posterior distributions have a straightforward interpretation as stating how plausible different values of the model parameters are given the observed data. We report point estimates, based on the median of posterior samples, and uncertainty intervals, based on the quantiles of posterior samples, that enclose the 95% most plausible estimates. In addition, we report the posterior probability, based on the proportion of posterior samples below zero, that an estimated mean effect size is negative.

@@

Figure 1 shows a flow diagram illustrating our search strategy, study selection, and data collection. Our preregistered search strategy returned `r f_comma(counts$n_included[1])` unique records from electronic databases. Of these, we excluded `r f_comma(counts$n_excluded[2])` (`r f_comma(counts$p_excluded[2])`) ineligible records after screening titles, abstracts, and keywords. Of the remaining `r f_comma(counts$n_included[2])` records, we excluded `r f_comma(counts$n_excluded[3])` (`r f_comma(counts$p_excluded[3])`) ineligible records after reviewing full-text manuscripts. We supplemented these records with `r f_comma(with(read_csv("../records/results/unpublished-studies.csv"), n_distinct(id)))` unpublished studies solicited from researchers and `r f_comma(with(read_csv("../records/results/citing-studies.csv"), n_distinct(id)))` records that cited at least three relevant works. Of `r f_comma(with(read_csv("../records/results/unpublished-studies.csv"), n_distinct(id)) + with(read_csv("../records/results/citing-studies.csv"), n_distinct(id)) + counts$n_included[3])` eligible studies, we had to exclude `r f_comma(counts$n_excluded[4])` (`r f_comma(counts$p_excluded[4])`) studies for which we could not extract or impute any relevant effect size. At each stage, we also excluded studies that used the same data as another study. Our final sample comprised effect sizes from `r paste0("$J = ", with(dl, f_comma(n_distinct(id))), "$")` studies spanning `r paste0("$N = ", with(distinct(dl, id, sample, n), f_comma(sum(n))), "$")` participants in `r paste0("$I = ", f_comma(nrow(distinct(dl, id, sample))), "$")` samples.

Figure 2 provides a qualitative overview of the literature. Even though most participants came from studies conducted in India (`r glue('$N = {f_comma(with(distinct(dl, id, sample, n, country), sum(n[country == "India"])))}$')`) and the United States (`r glue('$N = {f_comma(with(distinct(dl, id, sample, n, country), sum(n[country == "USA"])))}$')`; Figure 2a), most samples were collected in North America (`r glue('$I = {f_comma(with(distinct(dl, id, sample, n, continent), length(sample[continent == "North America"])))}$')`) and Europe (`r glue('$I = {f_comma(with(distinct(dl, id, sample, n, continent), length(sample[continent == "Europe"])))}$')`), with few samples from South America and Africa (Figure 2b). Most samples focused on relative inequalities that resulted from long-term migration (`r glue('$I = {f_comma(with(distinct(dl, id, sample, study_setting), length(sample[study_setting == "Long-term migration"])))}$')`), slavery (`r glue('$I = {f_comma(with(distinct(dl, id, sample, study_setting), length(sample[study_setting == "Slavery"])))}$')`), colonization (`r glue('$I = {f_comma(with(distinct(dl, id, sample, study_setting), length(sample[study_setting == "Colonization"])))}$')`), and short-term migration (`r glue('$I = {f_comma(with(distinct(dl, id, sample, study_setting), length(sample[study_setting == "Short-term migration"])))}$')`). With few exceptions, studies used observational, cross-sectional survey designs. Samples sizes ranged from `r with(distinct(dl, id, sample, n), paste0(f_comma(min(n)), " to ", f_comma(max(n)), " ($\\textit{Mdn} = ", f_comma(median(n)), "$)"))`. Unlike most psychological research, 2 in 5 samples used probability or representative sampling. Of all samples, `r with(distinct(dl, id, sample, study_intention), f_prop2percent(mean(study_intention == "Yes"), digits = 0))` (`r glue('$I = {f_comma(with(distinct(dl, id, sample, n, study_intention), length(sample[study_intention == "Yes"])))}$')`) were collected with the intention to examine the effects of intergroup contact on the outcomes considered in this meta-analysis.^[Many studies examined acculturation processes [@berry_immigration_1997] and measured both intergroup contact and perceived injustice without the relationship between the two being of interest.] Together, these observations highlight both limitations and strengths of the empirical literature.

@@

***Perceived Injustice.*** Across `r unique(pi_results$N)` participants from `r unique(pi_results$I)` samples in `r unique(pi_results$J)` studies, we found strong evidence for a weak association (`r pi_results$text[pi_results$parameter == "r_mean"]`) between intergroup contact and perceived injustice, with `r pi_results$p_below[pi_results$parameter == "r_mean"]` of posterior samples for the mean correlation coefficient falling below zero. We found evidence that correlation coefficients varied across studies (`r pi_results$text[pi_results$parameter == "tau_jj"]`) and across samples within studies  (`r pi_results$text[pi_results$parameter == "tau_ii"]`). Based on these analyses, we predicted that 80\% of studies would result in correlation coefficients between `r post_pred$r_pred_l80[post_pred$y_var == "pi"]` and `r post_pred$r_pred_u80[post_pred$y_var == "pi"]` and that researchers would need sample sizes of at least `r post_pred$n_req[post_pred$y_var == "pi"]` participants to find significant associations ($\alpha = .05$, two-sided) in 80\% of their studies.^[Sample sizes are based on posterior predictions from the three models, which implied that, for 80\% of studies, the absolute correlation coefficient would be `r post_pred$r_abs[post_pred$y_var == "pi"]` for perceived injustice, `r post_pred$r_abs[post_pred$y_var == "ca"]` for collective action, and `r post_pred$r_abs[post_pred$y_var == "ps"]` for policy support.]

@@

We conducted two kinds of moderator analyses. For both analyses, we had an insufficient number of effect sizes to examine moderators for any outcomes except perceived injustice (for details, see SOM-R).

@@

There is an emerging consensus that intergroup contact has the 'ironic' effect of undermining support for social change in disadvantaged groups. We conducted a preregistered meta-analytic test of this effect across 96 studies with 137 samples of 211,360 disadvantaged-group members. We found that, based on the available evidence, the associations of intergroup contact with perceived injustice (`r with(summary, p_text[y_var == "pi"])`), collective action (`r with(summary, p_text[y_var == "ca"])`), and support for reparative policies (`r with(summary, p_text[y_var == "ps"])`) were, on average, much more likely to be negative than positive. Thus, our meta-analysis seems to support the emerging consensus. We use the rest of this section to argue why this conclusion might be premature.

@@

A combination of moderators explained about a third of the between-study variance in the association between intergroup contact and perceived injustice. We found that, on average, this association was negative only in studies of adults that measured contact *directly* by asking, for example, about the number of friends from the advantaged outgroup. This suggests that, as hypothesized, the 'ironic' effects result from direct contact and not from mere exposure or other contextual factors. We also found that, on average, effect sizes were greater for studies in post-colonial settings or on short-term migration than for studies in other contexts. Future research should systematically investigate variance in how contact affects support for social change across settings and cultures.

Third, the available evidence is almost entirely from cross-sectional, observational studies and thus consistent with alternative explanations for the observed associations. Support for social change could reduce intergroup contact rather than the other way around. For example, disadvantaged-group members involved in collective action might purposefully avoid forming friendships with advantaged-group members. Alternatively, the association between intergroup contact and support for social change could be spurious with both being caused by an unobserved confounder. For example, a disadvantaged-group member's socioeconomic status might both expose them to more advantaged-group members and reduce their perception of injustice. Future research should prioritize longitudinal studies to confirm the direction of the observed associations and (field) experiments to rule out confounding and other alternative explanations.^[Published research includes only three longitudinal studies [@koschate_when_2012; @reimer_intergroup_2017; @tropp_crossethnic_2012], all of which conflated within-person change and between-person stability [@hamaker_critique_2015], two experimental studies with a no-contact control condition [@becker_friend_2013; @droogendyk_renewed_2016], and two intervention studies [@reimer_building_2021; @shani_effect_2017].]

Further, the observed association between intergroup contact and support for social change could be confounded by an alternative predictor that *increases* support for social change. Supporting Reimer et al.'s [-@reimer_intergroup_2017] argument, we found that positive intergroup contact was not associated with support for social change after controlling for negative contact which, in turn, was associated with *greater* support for social change. We found mixed evidence for ingroup contact as an alternative explanation for the 'ironic' effects of intergroup contact [@sengupta_ingroup_2015] but note that few studies measured both ingroup and outgroup contact. Going forward, researchers should clarify which aspects of intergroup contact should theoretically affect support for social change—and include measures that allow testing both the hypothesized and competing explanations.

>>>>

---
title: "Meta-Analysis of the 'Ironic' Effects of Intergroup Contact"
header: "'IRONIC' EFFECTS META-ANALYSIS"
keywords: intergroup contact, meta-analysis, perceived discrimination, collective action, policy support, social change, social injustice 
bibliography: references.bib
csl: apa.csl
abstract: 
  Growing evidence suggests that intergroup contact, psychology’s most-researched paradigm for reducing prejudice, might have the ‘ironic’ effect of reducing support for social change in disadvantaged groups. We conducted a preregistered meta-analysis of this effect across `r with(dplyr::distinct(readRDS("../data/dl.rds"), sample, id, n), numform::f_comma(dplyr::n_distinct(id)))` studies with `r with(dplyr::distinct(readRDS("../data/dl.rds"), sample, id, n), numform::f_comma(length(sample)))` samples of `r with(dplyr::distinct(readRDS("../data/dl.rds"), sample, id, n), numform::f_comma(sum(n)))` disadvantaged-group members. As predicted, intergroup contact was, on average, associated with less perceived injustice (`r with(readRDS("../results/results_preregistered_analyses.rds"), paste0("$r = ", numform::f_num(median(r_mean[x_var == "ic" & y_var == "pi"]), digits = 2), "$"))`), collective action (`r with(readRDS("../results/results_preregistered_analyses.rds"), paste0("$r = ", numform::f_num(median(r_mean[x_var == "ic" & y_var == "ca"]), digits = 2), "$"))`), and support for reparative policies (`r with(readRDS("../results/results_preregistered_analyses.rds"), paste0("$r = ", numform::f_num(median(r_mean[x_var == "ic" & y_var == "ps"]), digits = 2), "$"))`). Moderator analyses found that associations were strongest in studies of adults that measured intergroup contact directly and examined short-term migration or post-colonial intergroup relations. After controlling for the positive association of negative contact with support for social change, positive contact was not associated with any outcome in studies that measured both predictors. We close by highlighting open questions about the relationship between intergroup contact and social change.
statement: 
  As societies are becoming more diverse, there are more opportunities for members of different groups to come in contact with each other. Psychologists disagree about what this means for overcoming persistent inequality between advantaged and disadvantaged groups. Many argue that intergroup contact reduces advantaged-group members' prejudice and, thereby, prevents interpersonal discrimination. Reducing prejudice, however, is rarely enough to overcome institutional and structural discrimination. Instead, social change often results from disadvantaged-group members mobilizing against social injustice (for example, in the 2020 Black Lives Matter protests). Critics argue that positive intergroup contact could create a false sense of equality and, thereby, stifle disadvantaged-group members' opposition to injustice. We present the first systematic review and meta-analysis of the growing number of studies examining how intergroup contact relates to support for social change in disadvantaged groups. Our findings help us understand whether, when, and how intergroup contact might hinder, rather than help, social change.
output:
  pdf_document:
    latex_engine: pdflatex
    keep_tex: yes
    template: apa-template.tex
  word_document:
    reference_docx: word-template.docx
---

@@

It is ironic that psychology’s most-researched paradigm for reducing prejudice—intergroup contact—should hinder, rather than help, social change. Hundreds of studies have confirmed that intergroup contact is associated with less negative feelings toward outgroup members [for a meta-analysis, see @pettigrew_meta-analytic_2006]. Most studies have focused on advantaged-group members' contact experiences with disadvantaged-group members [@tropp_relationships_2005], guided by the idea that reducing advantaged-group members' prejudice will prevent interpersonal discrimination. Reducing prejudice, however, is rarely enough to overcome institutional and structural discrimination. Instead, as the 2020 Black Lives Matter protests have shown, social change often requires political mobilization by the disadvantaged. Critics have argued that, by fostering harmony and reducing conflict, intergroup contact might distract disadvantaged-group members from the injustice they face and undermine their motivation for social change [@dixon_beyond_2005; @reicher_rethinking_2007; @brown_strategic_2003]. By reducing disadvantaged-group members' opposition to injustice, intergroup contact might thus have the 'ironic' [@saguy_irony_2009], 'sedative' [@cakal_investigation_2011], or 'paradoxical' [@dixon_paradox_2010] effect of hindering social change.

Supporting this argument, initial studies provided evidence that contact with advantaged-group members reduces disadvantaged-group members' perceptions of injustice [@saguy_irony_2009], collective action [@cakal_investigation_2011], and support for reparative policies [@dixon_intergroup_2007]. Others, however, have pointed to conflicting evidence [@poore_contact_2002] and argued that intergroup contact might instead make discrimination more salient and, thereby, increase disadvantaged-group members' support for social change [@pettigrew_recent_2011]. In the decade since the initial studies, many more studies have examined the relationships between intergroup contact and support for social change in disadvantaged groups. We present a systematic review and meta-analysis of this growing literature. Specifically, we evaluate the evidence for three hypotheses about how intergroup contact might reduce support for social change.

Perceived injustice—that is, perceiving unjust group-based deprivation or discrimination against a disadvantaged ingroup—is a prerequisite to instigating social change [@van_zomeren_toward_2008]. Researchers have argued that positive contact with advantaged groups could contradict perceptions of personal discrimination and, thereby, make group discrimination appear less plausible [@dixon_paradox_2010]. Positive contact could further reduce perceived injustice by emphasizing commonalities over differences [@saguy_irony_2009] and by motivating disadvantaged-group members to adopt system-justifying ideologies [@sengupta_perpetuating_2013]. To the extent that it reduces perceived injustice, intergroup contact reduces support for social change.

Collective action—that is, any action directed at improving the conditions of the disadvantaged ingroup [@wright_responding_1990]—is crucial to achieving social change because advantaged-group members rarely surrender their advantages without sustained pressure from disadvantaged-group members [@blumer_race_1958; for a review, see @dixon_beyond_2012]. Researchers have argued that intergroup contact could discourage disadvantaged-group members from engaging in collective action by diminishing their perceived discrimination [@tropp_crossethnic_2012], by blurring boundaries between the disadvantaged ingroup and the advantaged outgroup [@saguy_irony_2009], by reducing negative attitudes toward the advantaged outgroup [@wright_struggle_2009], and by quelling their anger  [@hayward_how_2018; @tausch_how_2015]. To the extent that it discourages collective action, intergroup contact reduces support for social change.

Reparative policies (e.g., affirmative action) can be effective means to reduce social inequality. Disadvantaged-group members, who stand to benefit from their implementation, are important advocates for these policies. Researchers have argued that intergroup contact could diminish disadvantaged-group members’ support for redistributive policies by creating sympathy for advantaged-group members [@dixon_intergroup_2007], by fostering positive outgroup attitudes and reducing attention to inequality [@saguy_irony_2009], and by increasing their endorsement of system-justifying ideologies [@sengupta_perpetuating_2013]. To the extent that it diminishes support for redistributive policies, intergroup contact reduces support for social change.

Our first objective was to systematically review the available evidence and to evaluate its strengths and limitations. Our second objective was to synthesize the available evidence to establish the direction and magnitude of the relationships between intergroup contact and the three outcomes. Doing so allowed us to determine whether the available evidence supports a negative or positive [see @pettigrew_recent_2011] relationship between contact and support for social change and to estimate the strength of this relationship. Our third objective was to explore the variability of this relationship across study settings, designs, and other potential moderators. For example, we examined whether any effects were confined to Western, Educated, Industrialized, Rich, Democratic (WEIRD) countries, the cultural context of most psychological research [@henrich_weirdest_2010]. Our fourth objective was to assess how robust the available evidence was to publication bias.

In addition, we examined the evidence for two alternative accounts of this relationship. First, we considered studies that measured both ingroup and outgroup contact as predictors of the three outcomes. @sengupta_ingroup_2015 argued that contact with other members of the disadvantaged ingroup would increase political mobilization. As outgroup contact places limits on how much ingroup contact disadvantaged-group members can have [@pfister_contact_2020], the mobilizing effect of ingroup contact is an alternative explanation for negative associations between outgroup contact and support for social change.

Second, we considered studies that measured both negative and positive (outgroup) contact as predictors of the three outcomes. Researchers [@hayward_how_2018; @reimer_intergroup_2017] found negative contact with advantaged-group members to be associated with greater perceived discrimination and stronger collective action intentions. @reimer_intergroup_2017 argued that, to the extent disadvantaged-group members who have more positive contact also tend to have less negative contact, the mobilizing effect of negative contact is an alternative explanation for negative associations between outgroup contact and support for social change.

@@

We included studies if they measured or manipulated the quantity of, quality of, and opportunity for contact with members of outgroups that are relatively advantaged compared to the participants’ ingroup. We considered opportunity for contact, as it is a potential precursor to and proxy for face-to-face contact, but not imagined or extended contact. Similarly, we included studies that measured intergroup contact indirectly by, for example, asking what proportion of someone’s friends were *not* from the participants’ ingroup.

@@

***Perceived Injustice.*** We included studies that measured perceptions that one is discriminated against because of one’s group membership, that one’s group faces discrimination in society, that one’s group is relatively deprived compared to other groups, or that the deprivation and/or discrimination faced by one’s group is unjust and illegitimate. We also considered expectations of fair treatment as the reverse of perceived injustice. We included studies that measured personal or group discrimination or a mixture of both.

@@

Bayesian inference involves choosing a likelihood function and prior distributions. A likelihood function links the observed data to one or more model parameters and states how likely the observed data are given different values of said model parameters. Prior distributions state how plausible different values of said model parameters are before considering the observed data. Bayesian inference applies Bayes’ theorem to update prior distributions in light of the observed data to produce posterior distributions. Other than *p*-values and confidence intervals, the resulting posterior distributions have a straightforward interpretation as stating how plausible different values of the model parameters are given the observed data. We report point estimates, based on the median of posterior samples, and uncertainty intervals, based on the quantiles of posterior samples, that enclose the 95% most plausible estimates. In addition, we report the posterior probability, based on the proportion of posterior samples below zero, that the mean effect size is negative.

@@

Figure 1 shows a flow diagram illustrating our search strategy, study selection, and data collection. Our preregistered search strategy returned `r f_comma(counts$n_included[1])` unique records from electronic databases. Of these, we excluded `r f_comma(counts$n_excluded[2])` (`r f_comma(counts$p_excluded[2])`) ineligible records after screening titles, abstracts, and keywords. Of the remaining `r f_comma(counts$n_included[2])` records, we excluded `r f_comma(counts$n_excluded[3])` (`r f_comma(counts$p_excluded[3])`) ineligible records after reviewing full-text manuscripts. We supplemented these records with `r f_comma(counts$unpublished[3])` unpublished studies and `r f_comma(counts$citing[3])` records that cited at least three relevant works. Of `r f_comma(with(counts[3,], database + citing + unpublished))` eligible studies, we had to exclude `r f_comma(counts$n_excluded[4])` (`r f_comma(counts$p_excluded[4])`) studies for which we could not extract or impute any relevant effect size. At each stage, we also excluded studies that used the same data as another study. Our final sample comprised effect sizes from `r paste0("$J = ", with(dl, f_comma(n_distinct(id))), "$")` studies spanning `r paste0("$N = ", with(distinct(dl, id, sample, n), f_comma(sum(n))), "$")` participants in `r paste0("$I = ", f_comma(nrow(distinct(dl, id, sample))), "$")` samples (for a complete list of studies, see SOM-U).

Figure 2 provides a qualitative overview of the literature. Even though most participants came from studies conducted in India (`r glue('$N = {f_comma(with(distinct(dl, id, sample, n, country), sum(n[country == "India"])))}$')`) and the United States (`r glue('$N = {f_comma(with(distinct(dl, id, sample, n, country), sum(n[country == "USA"])))}$')`; Figure 2a), most samples were collected in North America (`r glue('$I = {f_comma(with(distinct(dl, id, sample, n, continent), length(sample[continent == "North America"])))}$')`) and Europe (`r glue('$I = {f_comma(with(distinct(dl, id, sample, n, continent), length(sample[continent == "Europe"])))}$')`), with few samples from South America and Africa (Figure 2b). Most samples focused on relative inequalities that resulted from long-term migration (`r glue('$I = {f_comma(with(distinct(dl, id, sample, study_setting), length(sample[study_setting == "Long-term migration"])))}$')`), slavery (`r glue('$I = {f_comma(with(distinct(dl, id, sample, study_setting), length(sample[study_setting == "Slavery"])))}$')`), colonization (`r glue('$I = {f_comma(with(distinct(dl, id, sample, study_setting), length(sample[study_setting == "Colonization"])))}$')`), and short-term migration (`r glue('$I = {f_comma(with(distinct(dl, id, sample, study_setting), length(sample[study_setting == "Short-term migration"])))}$')`). With few exceptions, studies used observational, cross-sectional survey designs. Samples sizes ranged from `r with(distinct(dl, id, sample, n), paste0(f_comma(min(n)), " to ", f_comma(max(n)), " ($\\textit{Mdn} = ", f_comma(median(n)), "$)"))`. Unlike most psychological research, 2 in 5 samples used probability or representative sampling. Of all samples, `r with(distinct(dl, id, sample, study_intention), f_prop2percent(mean(study_intention == "Yes"), digits = 0))` (`r glue('$I = {f_comma(with(distinct(dl, id, sample, n, study_intention), length(sample[study_intention == "Yes"])))}$')`) were collected with the intention to examine the effects of intergroup contact on the outcomes considered in this meta-analysis. Many studies instead examined acculturation processes [@berry_immigration_1997] and measured intergroup contact and perceived injustice without hypothesizing that the former would influence the latter. Together, these observations highlight both limitations and strengths of the empirical literature.

@@

***Perceived Injustice.*** Across `r unique(pi_results$N)` participants from `r unique(pi_results$I)` samples in `r unique(pi_results$J)` studies, we found strong evidence for a weak association (`r pi_results$text[pi_results$parameter == "r_mean"]`) between intergroup contact and perceived injustice, with `r pi_results$p_below[pi_results$parameter == "r_mean"]` of posterior samples for the mean correlation coefficient falling below zero. We found evidence that correlation coefficients varied across studies (`r pi_results$text[pi_results$parameter == "tau_jj"]`) and across samples within studies  (`r pi_results$text[pi_results$parameter == "tau_ii"]`). Based on these analyses, we predicted that 80\% of studies would result in correlation coefficients between `r post_pred$r_pred_l80[post_pred$y_var == "pi"]` and `r post_pred$r_pred_u80[post_pred$y_var == "pi"]` and that researchers would need sample sizes of at least `r post_pred$n_req[post_pred$y_var == "pi"]` participants to find significant associations ($\alpha = .05$, two-sided) in 80\% of their studies.^[Sample sizes are based on posterior predictions which implied that, for 80\% of studies, the absolute correlation coefficient would be `r post_pred$r_abs[post_pred$y_var == "pi"]` for perceived injustice, `r post_pred$r_abs[post_pred$y_var == "ca"]` for collective action, and `r post_pred$r_abs[post_pred$y_var == "ps"]` for policy support.]

@@

We conducted two kinds of moderator analyses. For both analyses, we had an insufficient number of effect sizes to examine moderators for any outcome except perceived injustice (for details, see SOM-R).

@@

There is an emerging consensus that intergroup contact has the 'ironic' effect of undermining support for social change in disadvantaged groups. We conducted a preregistered meta-analytic test of this effect across 96 studies with 137 samples of 211,360 disadvantaged-group members. We found that, based on the available evidence, the associations of intergroup contact with perceived injustice (`r with(summary, p_text[y_var == "pi"])`), collective action (`r with(summary, p_text[y_var == "ca"])`), and support for reparative policies (`r with(summary, p_text[y_var == "ps"])`) were, on average, much more likely to be negative than positive. Thus, our meta-analysis seems to support the emerging consensus. However, our findings qualify this conclusion in several important ways.

@@

A combination of moderators explained about a third of the between-study variance in the association between intergroup contact and perceived injustice. We found that, on average, this association was negative only in studies of adults that measured contact *directly* (e.g., by asking about the number of friends from the advantaged outgroup). This suggests that, as hypothesized, the 'ironic' effects result from direct contact and not from other contextual factors. We also found that, on average, effect sizes were greater for studies in post-colonial settings or on short-term migration than for studies in other contexts. Future research should systematically investigate variance in how contact affects support for social change across settings and cultures.

Third, the available evidence is almost entirely from cross-sectional, observational studies and thus consistent with alternative explanations for the observed associations. Support for social change could reduce intergroup contact rather than the other way around. For example, disadvantaged-group members involved in collective action might avoid forming friendships with advantaged-group members. Alternatively, the observed associations could be spurious with both being caused by an unobserved confounder. For example, a disadvantaged-group member's socioeconomic status might both expose them to more advantaged-group members and reduce their perception of injustice. Future research should prioritize longitudinal studies to confirm the direction of the observed associations and (field) experiments to rule out confounding and other alternative explanations.^[Published research includes only three longitudinal studies [@koschate_when_2012; @reimer_intergroup_2017; @tropp_crossethnic_2012], all of which conflated within-person change and between-person stability [@hamaker_critique_2015], two experimental studies with a no-contact control condition [@becker_friend_2013; @droogendyk_renewed_2016], and two intervention studies [@reimer_building_2021; @shani_effect_2017].]

Fourth, the observed association between intergroup contact and support for social change could be confounded by an alternative predictor that *increases* support for social change. Supporting Reimer et al.'s [-@reimer_intergroup_2017] argument, we found that positive intergroup contact was not associated with support for social change after controlling for negative contact which, in turn, was associated with *greater* support for social change. We found mixed evidence for ingroup contact as an alternative explanation for the 'ironic' effects of intergroup contact [@sengupta_ingroup_2015] but note that few studies measured both ingroup and outgroup contact. Going forward, researchers should clarify which aspects of intergroup contact should theoretically affect support for social change—and include measures that allow testing both the hypothesized and competing explanations.
